Concurrent Programming in Harmony

### 2021-07-24 book/paper.tex
Added explanation for each of the statements

The expression \textbf{const} \texttt{N = 3} introduces a new constant
\texttt{N} with the value 3.  Evaluating \texttt{N} does not lead to
loading from a memory location.  The assignment can be overridden with
the \texttt{-c} flag: \texttt{harmony -cN=4} executes the model checker
with 4 assigned to \texttt{N} instead of 3.
Harmony also supports \textbf{const} \texttt{N, M = 3, 4}, which
assigns 3 to \texttt{N} and 4 to \texttt{M}.

@@

@@

@@

\section{\textbf{if ... elif ... else}}

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

The expression \textbf{const} \texttt{N = 3} introduces a new constant
\texttt{N} with the value 3.  Evaluating \texttt{N} does not lead to
loading from a memory location.  The assignment can be overridden with
the \texttt{-c} flag: \texttt{harmony -cN=4} executes the model checker
with 4 assigned to \texttt{N} instead of 3.
Harmony also supports \textbf{const} \texttt{N, M = 3, 4}, which
assigns 3 to \texttt{N} and 4 to \texttt{M}.
Harmony also has limited support for \emph{constant folding}.  For example,
\textbf{const} \texttt{N = 3 + 4} assigns value 7 to constant \texttt{N}.

@@

Harmony does not support \textbf{break} or \textbf{continue} statements.

@@

The statement \textbf{from} \texttt{x} \textbf{import} \texttt{a, b, ...}
imports module \texttt{x} and makes its constants $a, b, ...$ also constants
in the current module.  If a module is imported more than once, its code
is only included the first time.  The constants will typically be the
names of methods (program counter constants) within the module.

You can import all constants from a module \texttt{m}
(including program counter constants) using the statement
\textbf{from} \texttt{m} \textbf{import} \texttt{*}.
This, however, excludes constants that start with the character \_: those
are considered \emph{private} to the module.

@@

The statement \textbf{go} $c e$ starts a thread with context $c$ that
has executed a \textbf{stop} expression.  The \textbf{stop} expression
returns value $e$.
The same context can be started multiple times, allowing threads to
\emph{fork}.

\section{\textbf{if ... [elif ...]* [else]}}

Harmony supports \textbf{if} statements.
In its most basic form, \textbf{if} $c: S_1; S_2; ...$ evaluates $c$
and executes statements $S_1, S_2, ...$ if and only if boolean expression
$c$ evaluated to true.
Harmony checks that $c$ is either \texttt{False} or \texttt{True}---if neither
is the case, Harmony reports an error.
The statement \textbf{if} $c: S_1, S_2, ...$ \textbf{else}:
$T_1; T_2; ...$ is similar, but executes statements $T_1, T_2, ...$
if and only if $c$ evaluated to false.
You can think of \textbf{elif} $c:$ as shorthand for
\textbf{else}: \textbf{if} $c:$.

@@

The statement \textbf{import} $\mathtt{m}_1, \mathtt{m}_2, ...$ imports modules
$\mathtt{m}_1, \mathtt{m}_2, ...$ in that order.
If a module is imported more than once,
its code is only included the first time.
The constants (including method constants) and shared variables declared 
in that module can subsequently be referenced by prepending ``\texttt{m}.''.
For example, method \texttt{f()} in imported module \texttt{m} is invoked
by calling \texttt{m.f()}.  If you would prefer to invoke it simply as
\texttt{f()}, then you have to import using the statement
\textbf{from} \texttt{m} \textbf{import} \texttt{f}.

@@

The statement \textbf{invariant} $c$ declares that boolean expression $c$
is an invariant.  $c$ is only allowed to read shared variables and
is evaluated atomically after every state change.
If it evaluated to \texttt{False} Harmony reports and error.  Harmony
also reports an error if the expression evaluates to a value other
than \texttt{False} or \texttt{True}.

@@

You can introduce new local variables in a method using the
\texttt{let} expression.
The statement \textbf{let} $a = b: S_1; S_2, ...$ evaluates $b$, assigns
the result to local variable $a$, and evaluates statements
$S_1, S_2, ...$.  \textbf{let} supports pattern matching, so you can
write \texttt{let} $x, (y, z) = b: S_1; S_2, ...$.  This will only work
if $b$ is a tuple with two elements, the second of which also being a
tuple with two elements---if not, Harmony will report an error.
The variables may be updates in statements $S_1, S_2, ...$.

\textbf{let} statements may be nested, such as
\textbf{let} $a_1 = b_1$ \textbf{let} $a_2 = b_2: S_1; S_2; ...$.
Doing so can improve readability by reducing indentation
compared to writing them as separate statements.
Compare the following two examples:

\begin{code}
\begin{tabbing}
XXX\=XXX\=XXX\=XXX\kill
\>\textbf{let} $a = y$: \\
\>\>\textbf{let} $b = z$: \\
\>\>\>...
\end{tabbing}
\end{code}
\begin{code}
\begin{tabbing}
XXX\=XXX\=XXX\=XXX\kill
\>\textbf{let} $a = y$ \\
\>\textbf{let} $b = z$: \\
\>\>...
\end{tabbing}
\end{code}

@@

The \textbf{pass} statement does nothing.

@@

The statement \textbf{select} $x$ \textbf{in} $y: S_1; S_2; ...$
requires that $y$ evaluates to a set value.
The statement does the following three things atomically:
\begin{itemize}
\item it waits until $y$ is non-empty;
\item it selects one element of $y$ non-deterministically (using
a \texttt{choose} expression);
\item it executes statements $S_1, S_2, ...$ with the selected element
assigned to local variable $x$.
\end{itemize}
$x$ may be a pattern, like in \texttt{let}, \texttt{for}, and \texttt{def}
statements.
Harmony reports an error if $y$ evaluates to a value that is not a set.
If \textit{waiting} is an unused local variable,
then \textbf{select} $x$ \textbf{in} $y: S_1; S_2; ...$ is equivalent to

\begin{code}
\begin{tabbing}
XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\=XXX\kill
\>\textbf{let} $\mathit{waiting} = \mathtt{True}:$ \\
\>\>\textbf{while} \textit{waiting}: \\
\>\>\>\textbf{atomic}: \\
\>\>\>\>\textbf{if} $y$ != \{\}:\\
\>\>\>\>\>\textbf{let} $x = \mathbf{choose}(y)$:\\
\>\>\>\>\>\>$S_1$ \\
\>\>\>\>\>\>$S_2$ \\
\>\>\>\>\>\>$...$ \\
\>\>\>\>\>$\mathit{waiting} = \mathtt{False}$
\end{tabbing}
\end{code}

The statement is particularly useful in programming network protocols
when having to wait for one or more messages and executing a set
of actions atomically after the desired messages have arrived.

@@

In Harmony, shared variable \texttt{Load} and \texttt{Store} operations
are atomic and have \emph{sequential consistency}.
However, Harmony does check for \emph{data races}.  A data race occurs
when two or more threads simultaneously access the same shared variable,
with at least one of the accesses being a \texttt{Store} operation
\emph{outside of an atomic block}.
If so, Harmony will report an error.  This error can be suppressed by
declaring the shared variable as sequential.  In particular, the statement
\texttt{sequential} $x, y, ...$ specifies that the algorithm assumes
that the given variables have sequential consistency.

Note that few modern processors support sequentially consistent memory
by default, as doing so would lead to high overhead.

@@

The statement \texttt{spawn} $m$ $a$ starts a new thread that executes
method $m$ with argument $a$. $m$ must be a program counter constant,
and $a$ is typically a tuple containing
zero or more parameters to be passed to the method.

The default thread-local state of the thread, called \textit{self},
is the empty dictionary by default.  It can be specified by adding
a parameter: \texttt{spawn} $m$ $a, e$ specifies that $e$ should be
the initial value of the thread-local state.

@@

The statement \textbf{trap} $m$ $a$ specifies that the current thread
should execute method $m$ with argument $a$ and some future, unspecified,
time.  It models a timer interrupt or any kind of asynchronous event to
be handled by the thread.  Such interrupts can be disabled by setting
the interrupt level of the thread to \texttt{True} using the
\textbf{setintlevel} operator.

@@

The statement \textbf{while} $c: S_1; S_2; ...$ executes statements
$S_1, S_2, ...$ repeatedly as long as $c$ evaluates to \texttt{True}.
Harmony does not support \textbf{break} or \textbf{continue} statements.
